---
title: "Calculating confidence regions using the cube-root bootstrap method"
author: "Panaghis Mavrokefalos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating confidence regions using the cube-root bootstrap method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

For parameter and index definitions (e.g. `mIdx`, `noAttr`, `dataArray`), see `vignette("glossary", package = "maxscoreest")`.

The function `cubeRootBootstrapCR()` implements a cube-root bootstrap method
for calculating the confidence regions of the $\beta$ parameters that are
obtained by `optimizeScoreFunction()`. This approach is applicable to situations
in which the asymptotics is in the number of markets. See the
`pointIdentifiedCR()` function, and the `"matched"` vignette for more
information.

The implementation is based on the paper [1]. Links to the
[paper](https://mdcattaneo.github.io/papers/Cattaneo-Jansson-Nagasawa_2020_ECMA.pdf),
its [supplement](https://mdcattaneo.github.io/papers/Cattaneo-Jansson-Nagasawa_2020_ECMA--Supplement.pdf),
and an [implementation](https://github.com/mdcattaneo/replication-CJN_2020_ECMA) are provided.

## Overview

Let's set up the data array using the synthetic data, and compute the estimates
$\hat{\beta}$.

```{r setup}
library(maxscoreest)
filename <- system.file("extdata", "precomp_testdata.dat", package = "maxscoreest")
matchedData <- importMatched(filename)
ineqmembers <- Cineqmembers(matchedData$mate)
dataArray <- CdataArray(matchedData$distanceMatrices, ineqmembers)
bounds <- makeBounds(matchedData$noAttr, 10)
optimParams <- getDefaultOptimParams()
randomSeed <- 42
optimizeScoreArgs <- list(
    dataArray = dataArray,
    bounds = bounds,
    optimParams = optimParams,
    getIneqSat = TRUE,
    permuteInvariant = TRUE,
    numRuns = 1000)
set.seed(randomSeed)
optResult <- do.call(optimizeScoreFunction, optimizeScoreArgs)
pointEstimate <- optResult$optArg
pointEstimate
```

`cubeRootBootstrapCR()` has the same parameter list as `pointIdentifiedCR`,
except it is not necessary to specify the `ssSize` parameter, as the method
generates its bootstrap samples by drawing (with replacement) as many markets as
there are in the full data array.

```{r}
numSubsamples <- 20
ssSize <- NULL
confidenceLevel <- 0.95
groupIDs <- makeGroupIDs(ineqmembers)
```

Let's call the function with the default options:
```{r}
options <- list()
set.seed(randomSeed)
cr <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
cr
```

The member `cr$cr` is an array containing the *confidence regions*. Each
column is a pair of values denoting the confidence interval of the corresponding
attribute. The members `cr$rawEstimates` and `cr$estimates` are also arrays,
containing the results of the bootstrap operation. Each row corresponds to a
single bootstrap sample. The rows of `cr$rawEstimates` are the actual
parameter vectors $\tilde{\beta}$ obtained by optimizing the bootstrap function,
while the rows of `cr$estimates` are the vectors $\tilde{\beta} - \hat{\beta}$,
where $\hat{\beta}$ are the point estimates. We call the former estimates
*uncentered*, and the latter *centered*.

The matrix `cr$H` is the calculated estimate of the matrix $H$ mentioned in the
paper [1]. While its value can be ignored in the usual case, it can allow for a
more sophisticated use of the package. The significance of this matrix is
explained in the paper, while the final section of this document gives an
overview of the method.

## More on the method

The estimates are calculated by sampling, with replacement, a number of markets
equal to their total number in the original data, creating its associated data
array, and maximizing a function related to the difference of their scores,
augmented by a quadratic term.

In particular, let $S(\beta; X)$ be the score function corresponding to the data
array $X$ (see also `makeScoreObjFun()`), $\hat{\beta}$ be the point estimate
previously calculated, and $H$ (short for $\tilde{H}_n$) be the matrix which
estimates $H_0$. After sampling the full data array $X_{\mathrm{full}}$,
we create the sample data array $X_{\mathrm{sample}}$.

Define now the quadratic term
$q(\beta; \hat{\beta}, H) = \frac{1}{2} (\beta - \hat{\beta})^T H (\beta - \hat{\beta})$

Then the function maximized at each bootstrap step is
$B(\beta; \hat{\beta}, H, X_{\mathrm{full}}, X_{\mathrm{sample}}) = c S(\beta; X_{\mathrm{sample}}) - S(\beta; X_{\mathrm{full}}) - q(\beta; \hat{\beta}, H)$

Note that, as defined, the score function is normalized on the number of
inequalities of the data array, and therefore takes values in the interval
$[0, 1]$.

The *correction factor* $c$ has a value which, by default, is equal to
the ratio of the number of inequalities in the sample data array ($m$) to the
number of inequalities in the full data array ($n$). Since markets are, in the
general case, unbalanced with respect to their size, this factor is not
necessarily equal to $1$. In a previous version, this factor was fixed to $1$,
but most users should not revert to this behavior unless reproducibility with
that version is desired.

## Parameters/Options

As in `pointIdentifiedCR()`, a list of arguments related to optimizing
the above function is required. The optimization is performed by the
function `optimizeBootstrapFunction()`, which is similar to
`optimizeScoreFunction()`. The parameter `optimizeScoreArgs`
should be a list with the following elements: 

- `bounds # of the (coefficient) estimates`
- `coefficient1`
- `method`
- `optimParams`

The list of arguments to `optimizeBootstrapFunction()` is then internally
constructed using the above data.

`cubeRootBootstrapCR()` supports passing an `options` list. The relevant options
are explained here, but the user can also consult its documentation for more
details.

### `centered`

Whether to *center* (see above, regarding `$rawEstimates` and `$estimates`) the
confidence regions `$cr`. This can be useful if the user desires that the
confidence intervals share a common reference.

The rest of the options are related to the estimation of $H$. See also the
`makeHmatrix()` function.

### `Hest`

Selects the method for estimation. Set to `"numder"` for the numerical
derivative method, which is more general, or to `"plugin"`, for an
estimator which is specific to this problem. We recommend using the plug-in
method, which is the default.

The plug-in method assumes a kernel function with specific properties.
Specifically, its antiderivative should approximate the indicator function, and
Condition K from the supplement of paper [1] should be satisfied. The only
kernel currently available in the code is the function $K(u) = \phi(u)$, i.e.
the pdf of the standard normal distribution.

### `bw`

Both the numerical derivative and the plug-in method depend on a matrix-valued
parameter: the *step* $\epsilon_{n,kl}$, and the *bandwidth* $h_{n,kl}$
respectively. This is supposed to be a matrix of size $d \times d$, where $d$
is the number of *free* attributes (i.e. the total number of attributes
reduced by one, or the length of the vector $\beta$), but if given as a scalar,
it is automatically converted to a matrix of the appropriate size with constant
elements. Its default value is the scalar `1`.

The paper [1] provides an alternative method of calculating this parameter,
which is named the *Rule-Of-Thumb* (ROT) method. Setting the option `bw` to
`"rot"` allows the step or bandwidth to be automatically calculated.

### `makePosDef` and `makePosDefTol`

The matrix $H$ is, by definition, positive semidefinite. In practice, it is
possible, though, that the estimate is not positive semidefinite. Setting
`makePosDef` to `TRUE` replaces the initial estimate with a positive
semidefinite matrix which is closest, in a sense, to that estimate. We provide
two approaches:

- Adding a constant value $\kappa$ to each diagonal element of $H$. This value
  is defined as $\epsilon_{\mathrm{tol}} - \lambda_d$, where $\lambda_d < 0$ is
  the smallest eigenvalue of the $H$ estimate, and $\epsilon_{\mathrm{tol}}$ is
  a non-negative tolerance. This tolerance is specified by the value of
  `makePosDefTol`.
 
- Replacing all negative eigenvalues with $0$. To select this approach, set
  `makePosDefTol` to `"drop"`.

### `useCorrectionFactor`

---

**Warning!**
We strongly recommend not changing this option unless the user has a reason to
check the results for a different formula for the correction factor.

---

Selects whether the constant factor $c$ in the bootstrap function should equal
$\frac{m}{n}$ (`TRUE`, the default) or revert to the legacy behavior, where it
used to equal $1$ (`FALSE`). See the section "More on the method" for the
definition of this factor.

### `Hbypass`

Although the matrix $H$ is automatically estimated by the package in the usual
case, we allow the users to skip this procedure and provide their own estimate
instead. This is achieved by setting the option `Hbypass` to the provided
estimate. Note that the user is responsible for ensuring that the estimate is a
positive semidefinite matrix of the correct dimensions.

## Examples

### Compare plug-in and numerical derivative methods

```{r compare1}
options$bw <- 0.8
options$Hest <- "numder"
set.seed(randomSeed)
crNumder <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crNumder[c("cr", "H")]

options$Hest <- "plugin"
set.seed(randomSeed)
crPlugin <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crPlugin[c("cr", "H")]
```

### Use ROT
```{r rot}
options$bw <- "rot"
options$Hest <- "numder"
set.seed(randomSeed)
crNumder <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crNumder[c("cr", "H")]

options$Hest <- "plugin"
set.seed(randomSeed)
crPlugin <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crPlugin[c("cr", "H")]
```

### Fix non-PSD $H$ matrix

```{r makePosDef}
pointEstimate <- optResult$bestRuns[[67]]$optArg
pointEstimate
options$makePosDef <- FALSE
set.seed(randomSeed)
crNonPSD <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crNonPSD[c("cr", "H")]
eigen(crNonPSD$H)$values # has a negative eigenvalue, is not PSD

options$makePosDef <- TRUE
set.seed(randomSeed)
crMakePosDef <- cubeRootBootstrapCR(
    dataArray, groupIDs, pointEstimate,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs, options)
crMakePosDef[c("cr", "H")]
eigen(crMakePosDef$H)$values
```

## References
- [1] M. D. Cattaneo, M. Jansson, and K. Nagasawa, “Bootstrap-Based Inference for Cube Root Asymptotics”, *Econometrica*, vol. 88, no. 5, pp. 2203–2219, September 2020.
