---
title: "Calculating optimal parameters from matched data"
author: "Panaghis Mavrokefalos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating optimal parameters from matched data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A quick example

For parameter and index definitions (e.g. `mIdx`, `noAttr`, `dataArray`), see `vignette("glossary", package = "maxscoreest")`.

```{r setup}
library(maxscoreest)
```

First, we need to import the data into R. Let's use some synthetic data provided
with the package.

```{r}
filename <- system.file("extdata", "precomp_testdata.dat", package = "maxscoreest")
```

We import this using the function `importMatched`.

```{r}
matchedData <- importMatched(filename)
```

Next, we need to calculate the *inequality members*, which are lists of indices
of the pairings that participate in the inequalities defining the objective
function. The function `Cineqmembers` uses the matching information, contained
in the list member `$mate`, to create these indices.

```{r}
ineqmembers <- Cineqmembers(matchedData$mate)
```

Now we can use the inequalities to create the *data array*. The function
`CdataArray` uses the indices of the inequality members to gather and combine
values from the distance matrices. Here, a *distance* is an attribute that
characterizes a pair composed of upstream and downstream agents (e.g. geographic
distance between the agents, or their age multiplied, etc.).

```{r}
dataArray <- CdataArray(matchedData$distanceMatrices, ineqmembers)
```

Next, we need to set up the optimization procedure.

---

## An important note regarding optimization

We are optimizing a non-smooth, non-convex objective function, so the
general-purpose optimization methods are ill-suited for this problem. We are
currently supporting only the *Differential Evolution* method (implemented in
the package `DEoptim`), but this is a **non-deterministic** method. A user can
fix the RNG seed on their setup, but there is no guarantee of reproducibility
across different machines, operating systems, R versions, or package versions.

These two facts stated above complicate the optimization in the following ways:

- The objective function is non-convex: this means that there is no guarantee of
reaching the global optimum using local search methods.
- The objective function is non-smooth, and actually piecewise constant: this
means that the objective attains its (global) optimum for infinitely many
points. Even worse, there are potentially many disconnected subsets of the
search space on which the objective attains its optimum.

Differential Evolution works very well in practice however, either finding the
global optimum or approximating it very closely.

---

Let's use the default parameters for Differential Evolution, and select a box
constraint for the paramater estimates of $[-100, 100]$. The user needs to
decide whether to restrict the parameter estimates to some bounds, whether the
bounds are symmetrical, etc. See below under "Optimization" a discussion of
these issues.

```{r}
bounds <- makeBounds(matchedData$noAttr, 100)
optimParams <- getDefaultOptimParams()
```

We might also want to fix the RNG for reproducibility across runs. This is
highly recommended. See the *note regarding optimization* above.

```{r}
randomSeed <- 42
set.seed(randomSeed)
```

We are now ready to perform the optimization. The function
`optimizeScoreFunction` creates and optimizes the objective.
Note that we don't have to pass the arguments via `do.call`, but since we'll
need the argument list later when calculating the confidence region, this way is
more convenient.

```{r}
optimizeScoreArgs <- list(
    dataArray = dataArray,
    bounds = bounds,
    optimParams = optimParams,
    getIneqSat = TRUE,
    permuteInvariant = TRUE,
    numRuns = 1,
    progressUpdate = 0)
optResult <- do.call(optimizeScoreFunction, optimizeScoreArgs)
```

The rest of the (optional) arguments are explained below:

- `getIneqSat`: Whether to include the boolean vector of satisfied inequalities
`$ineqSat` in the result.
- `permuteInvariant`: Whether to reorder the attributes rows of the data array
before and after the optimization, such that the attribute whose row has the
smallest standard deviation comes first. Note that setting this to `TRUE`
affects only the internal procedures, as the initial order is restored in the
results. In addition to specifying the RNG, this feature is highly recommended
in order to increase the likelihood that results are reproducible across runs
and computers.
- `numRuns`: Setting this to an integer greater than `1` restarts the
optimization method as many times as indicated by this integer. This is highly
useful for methods such as Differential Evolution, because such methods can
yield a different result depending on the random seed. Thus, this approach
increases the likelihood that the user searches and finds the estimates that
maximize a higher number of inequalities. The user can retrieve the runs with
the best overall value from the member `$bestRuns` in the result. See the
section "Restarting the optimization multiple times" below for more information.
- `progressUpdate`: If greater than `0`, prints a progress update to the console
for the specified increment in the number of completed runs of the optimization
procedure (e.g. every 8 runs, if the progress update is equal to `8`).

We can now see the optimal parameter vector, and how many inequalities are
satisfied across each market.

```{r}
optResult$optArg
optResult$optVal
calcPerMarketStats(optResult$ineqSat, makeGroupIDs(ineqmembers))
```

A full runnable script for this workflow is available as a workbook; see the README (Workbook templates).

# Calculating the confidence intervals

Now that we have an optimal solution for the parameters, we can use it to
calculate an estimate for the confidence regions. The procedure for constructing
the confidence intervals is explained in detail by Fox (2018) -- see both the
text and the Supplementary Material. The procedure is based on Romano and Shaikh
(2008).

To implement it, one needs to choose a number of iterations for the confidence
regions calculation procedure (`numSubsamples`). In addition, if the asymptotics
is in the number of markets (or nests), one needs to also choose the size of the
market subsets of each iteration (`ssSize`). Alternatively, the asymptotics
could be in the number of matches, but this approach is not implemented here.
Fox (2018) provides an example in which confidence intervals are generated by
subsampling from the number of matches within markets.

```{r}
ssSize <- 2
numSubsamples <- 50
```

We also need to map each inequality to its market, using the function
`makeGroupIDs`.

```{r}
groupIDs <- makeGroupIDs(ineqmembers)
```

Now we are ready to calculate the confidence regions. We chose the standard
value for the confidence level equal to $0.95$, but this can be modified to any
desired value.

```{r}
confidenceLevel <- 0.95
cr <- pointIdentifiedCR(
    dataArray, groupIDs, optResult$optArg,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs)
```

We can use the function `plotCR` to visualize the confidence region estimates
(centered around zero).

```{r}
plotCR(cr$estimates)
```

We can also see which market subsets were chosen at each iteration.

```{r}
cr$samples
```

A novel feature of this code is that instead of the subsampling method for
generating the confidence intervals proposed by Fox (2018), it is also possible
to use the cube-root bootstrap method introduced by Cattaneo, Jansson and
Nagasawa (2020). The signatures are the same:

```{r}
crCubeRoot <- cubeRootBootstrapCR(
    dataArray, groupIDs, optResult$optArg,
    ssSize, numSubsamples, confidenceLevel,
    optimizeScoreArgs)
```

See the vignette *Calculating confidence regions using the cube-root bootstrap method* (accessed by `vignette("cubeRootBootstrap", package = "maxscoreest")`) for more details on this method.

# Taking a closer look

## Importing

The `importMatched` function can handle delimiter-separated value files with
headers. It supports two modes:

- identifying fields by their position (the default)
- identifying fields by their name

The optional argument `fieldMode` switches between those modes.
The exact format of the header and values is explained in more detail in the
corresponding documentation.

## Accessing the imported data

We can see how many markets are in the data, how many members each stream
has in each market, and how many attributes there are.

```{r}
matchedData$noM
matchedData$noU
matchedData$noD
matchedData$noAttr
```

We can also view the file header, containing the field names.

```{r}
matchedData$header
```

The `matchedData` structure exposes two elements, called `$distanceMatrices` and
`$matchMatrices`. They are lists of size `noM` each, with each element
containing an array with the relevant data for each market:

```{r}
length(matchedData$distanceMatrices)
length(matchedData$matchMatrices)
```

Let's check whether, in the third market, there is a match between the
upstream-downstream pair indexed by $(4, 7)$. Note the order in which the
indices appear: market, downstream and then upstream. A value of `1` indicates a
match and a value of `0` indicates the respective upstream-downstream pairing is
not a match in the real data.

```{r}
mIdx <- 3
uIdx <- 4
dIdx <- 7
matchedData$matchMatrices[[mIdx]][dIdx, uIdx]
```

We can also view the matching information for a single stream (upstream or
downstream agents).

```{r}
# Leaving an index blank takes all elements for that dimension.
matchedData$matchMatrices[[mIdx]][, uIdx]
# `which` returns the indices where the elements above are `1` -- i.e. the
# downstream matches of upstream '4'.
which(matchedData$matchMatrices[[mIdx]][, uIdx] == 1)
```

We can view the distance attributes for a single attribute index, or for the
entire attribute range.

```{r}
aIdx <- 3
matchedData$distanceMatrices[[mIdx]][aIdx, dIdx, uIdx]
matchedData$distanceMatrices[[mIdx]][, dIdx, uIdx]
```

## Accessing the inequality members

The structure `ineqmembers` is a list of length `noM`. Each element contains
the indices which define each inequality for that market, packaged in a list.

```{r}
mIdx <- 1
length(ineqmembers)
names(ineqmembers[[mIdx]])
```

The `numIneqs` element is the number of inequalities for that market. The
other four elements are lists, with length equal to `numIneqs`.

```{r}
ineqmembers[[mIdx]]$numIneqs
length(ineqmembers[[mIdx]]$fctUpIdxs)
```

Each inequality, when everything has been moved to the LHS, contains terms of
the form $f_{\beta}(m, p, q) - f_{\beta}(m, r, s)$, where $f_{\beta}$ is the
payoff function of the match evaluated at the parameter vector $\beta$, and
$f_{\beta}(m, i, j)$ is its value for the upstream-downstream match $(i, j)$ of
the $m$-th market. The indices $p, q, r, s$ are stored the vectors
`$fctUpIdxs`, `$fctDnIdxs`, `$cfcUpIdxs`, and `$cfcDnIdxs`, respectively,
where `fct` and `cfc` correspond to the factual/LHS $(p, q)$ and
counterfactual/RHS $(r, s)$ indices, and `Up` and `Dn` to the upstream
$(p, r)$ and downstream $(q, s)$ indices.

For example, let's choose the first term of the first inequality for the
previous match:

```{r}
mIdx <- 3
uIdx <- 4
dIdx <- 7
ineqIdx <- 1
termIdx <- 1
ineqmembers[[mIdx]]$fctUpIdxs[[ineqIdx]][termIdx]
ineqmembers[[mIdx]]$fctDnIdxs[[ineqIdx]][termIdx]
ineqmembers[[mIdx]]$cfcUpIdxs[[ineqIdx]][termIdx]
ineqmembers[[mIdx]]$cfcDnIdxs[[ineqIdx]][termIdx]
```

Since the above expressions are cumbersome, we provide the `getIneqTermsIdxs`
and `getAllIneqTermsIdxs` functions, which provide the above information in a
more user-friendly mode. `getIneqTermsIdxs` returns an array with all indices
of a single inequality. `getAllIneqTermsIdxs` does the same for an entire
market.

```{r}
getIneqTermsIdxs(ineqmembers[[mIdx]], ineqIdx)
getAllIneqTermsIdxs(ineqmembers[[mIdx]])[[ineqIdx]]
```

## Accessing the data array

The `dataArray` object is an array with precomputed values used in calculating
the value of the objective function. Its rows correspond to attributes, and its
columns to inequalities.

```{r}
dim(dataArray)
```

The `dataArray` is "flat", in the sense that there is no distinction between
markets. To recover that distinction, we need to create a mapping from its
column indices to the market indices. This is handled by the function
`makeGroupIDs`

```{r}
groupIDs
```

We can now choose the columns corresponding to the second market...

```{r}
mIdx <- 2
qualifiedIndices <- which(groupIDs == mIdx)
qualifiedIndices
dataArray[, qualifiedIndices]
```

...or, more generally, to a subset of market indices:

```{r}
selectedGroups <- c(1, 3)
qualifiedIndices <- which(groupIDs %in% selectedGroups)
qualifiedIndices
```

## Optimization

The function `makeBounds` creates equal symmetrical box constraints for each
variable estimate:

```{r}
makeBounds(matchedData$noAttr, 100)
```

If non-symmetrical constraints for the estimates are desired, we can pass a
third argument to the function:

```{r}
makeBounds(matchedData$noAttr, 10, -20)
```

Otherwise, we can also create custom bounds. That object should be a list
with elements named `$lower` and `$upper`. Each element is a vector defining
lower and upper bounds for each variable.

```{r}
bounds <- list(lower=c(-1, -2, -3, -5), upper=c(2, 3, 5, 8))
bounds
```

Note that since we have 5 distance attributes, the number of free parameters
is 4.

`getDefaultOptimParams` returns a reasonable collection of parameters for the
Differential Evolution method. Therefore, the user should impose constraints on
the parameter estimates only if they have reasons to suspect the data is of a
particular form that requires such constraints.

```{r}
optimParams
```

See the documentation of `DEoptim::DEoptim.control` for more information and how
to fine-tune the parameters of the Differential Evolution method to obtain
better results for a given data.

To see the original names of the distance attributes for the optimal parameter
vector, one can use the following commands:

```{r}
distNames <- matchedData$header[matchedData$colIdxs$distanceColIdxs]
# Drop first value.
distNamesFree <- distNames[-1]
names(optResult$optArg) <- distNamesFree
optResult$optArg
```

### Restarting the optimization multiple times

Since we are dealing with an optimization problem with multiple solutions, we
provide an option to restart optimization procedures, such as *Differential
Evolution*, that depend on the state of the random generator. By using a
different random state on each run, the optimal parameter vectors calculated
on each run will also vary. The user can then choose one or more solutions
according to their own criteria. The overall goal is to reach the maximum number
of inequalities predicted.

Setting the parameter `numRuns` in `optimizeScoreFunction()` to an integer
greater than `1` will restart the optimization procedure that many times. The
list `bestRuns` then collects the results that share the overall best score
value (that is, it collects the solution(s) that provide(s) the highest number
of inequalities predicted).

It might be helpful to set the `progressUpdate` parameter to a positive value,
in order to track the progress of the long-running operation.

```{r restartOpt}
optimizeScoreArgs$bounds <- bounds <- makeBounds(matchedData$noAttr, 10)
optimizeScoreArgs$numRuns <- 1000
optimizeScoreArgs$progressUpdate <- 50
optResult <- do.call(optimizeScoreFunction, optimizeScoreArgs)
# How many runs, out of the initial 1000, gave the best overall score?
length(optResult$bestRuns)
# Show the result of the 100th such run.
optResult$bestRuns[[100]]
```

## References
- [1] Fox, J.T. (2018). Estimating matching games with transfers. *Quantitative
Economics*, 9: 1-38. <https://doi.org/10.3982/QE823>
- [2] Romano, J. P. and A. M. Shaikh (2008). Inference for identifiable
parameters in partially identified econometric models. *Journal of Statistical
Planning and Inference*, 138 (9), 2786–2807.
<https://doi.org/10.1016/j.jspi.2008.03.015>
