---
title: "Modifying markets: unmatching, removing, and rematching agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying markets: unmatching, removing, and rematching agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

After importing a matched dataset and (optionally) estimating the parameters,
you may want to **edit** one or more markets: break a specific match, remove
an agent altogether, and possibly re-solve the assignment problem afterwards.
The `modifydata` helpers provide a small, composable interface for doing this.

This is the R counterpart of the Mathematica `modifydata.m` module. Because R
passes data by value, the helpers operate on an explicit **market state list**
rather than on a fixed set of globals. The state is just the list returned by
`importMatched()`, optionally augmented with per-market quota vectors and
(for rematching) evaluated payoff matrices.

The functions covered in this vignette are:

| Function | Role |
| --- | --- |
| `storeState`, `restoreState` | Deep-copy snapshots of the state |
| `mateFromMatchMatrix`, `mateFromMatchMatrices` | Rebuild a mate `data.table` from a match matrix |
| `quotaFromMatchMatrix`, `addQuotasFromMatches` | Derive per-agent match counts |
| `modifyMarket` | Main editor — unmatch, remove, and/or rematch a single market |
| `removeUpstreams`, `removeDownstreams`, `removeStreams` | Lightweight slicing helpers |

## Setup

```{r setup}
library(maxscoreest)
```

We use the small synthetic dataset shipped with the package
(`precomp_proof.dat`, three markets) so the matrices are short enough to print
in full.

```{r}
filename <- system.file("extdata", "precomp_proof.dat", package = "maxscoreest")
state <- importMatched(filename)
```

```{r}
state$noM
state$noU
state$noD
state$noAttr
```

We have three markets with `(noU, noD) = (3, 2), (2, 3), (4, 4)` and three
distance attributes. Let's look at the initial match matrix for market 3:

```{r}
state$matchMatrices[[3]]
```

The convention used throughout the package is `matchMatrices[[m]][d, u]` —
rows are downstream agents, columns are upstream agents. The matrix above
shows for example that in market 3, upstream 1 is matched with downstreams 1
and 2.

## A quick example

Suppose we want to unmatch the pair (upstream 1, downstream 1) in market 3,
and see the resulting market state.

```{r}
s <- modifyMarket(state, m = 3, u = 1L, d = 1L,
                  unmatch = TRUE, remove = FALSE)
s$matchMatrices[[3]]
```

The cell `[1, 1]` flipped from 1 to 0. The original `state` is unchanged:

```{r}
state$matchMatrices[[3]][1, 1]
```

This is the most important convention to remember: **`modifyMarket` and the
convenience helpers always return a new state and never mutate the caller's
data**, including the reference-typed `mate` `data.table` objects.

## Adding quotas to the state

`modifyMarket` can update or use per-market quota vectors, but these aren't
part of the value returned by `importMatched()`. The helper
`addQuotasFromMatches()` derives them from the current match matrices (column
sums for upstreams, row sums for downstreams):

```{r}
state <- addQuotasFromMatches(state)
state$quotasU
state$quotasD
```

These are needed whenever you set any `quotaUpdate*`, `quotaReset`, or
`rematch` flag (the function errors with a clear message if you forget).

## Operation 1: unmatching specific pairs

Unmatching means clearing one or more `(d, u)` cells of the match matrix
without removing any agents. Pass aligned vectors `u` and `d` of equal length:

```{r}
s <- modifyMarket(state, m = 3, u = c(1L, 4L), d = c(2L, 4L),
                  unmatch = TRUE, remove = FALSE)
state$matchMatrices[[3]]    # before
s$matchMatrices[[3]]        # after: [2, 1] and [4, 4] are now 0
```

The mate table is automatically rebuilt so it stays consistent with the new
match matrix:

```{r}
s$mate[[3]]
```

A warning is emitted if you try to unmatch a pair that was already
unmatched — the operation still proceeds (it's a no-op in that case), but the
warning helps catch mis-specified inputs.

### Updating quotas on unmatch

By default, an unmatch does **not** touch the quota vectors. Setting
`quotaUpdateUpstream = TRUE` and/or `quotaUpdateDownstream = TRUE` recomputes
the corresponding side from the new match matrix:

```{r}
s <- modifyMarket(state, m = 3, u = 1L, d = 1L,
                  unmatch = TRUE, remove = FALSE,
                  quotaUpdateUpstream = TRUE,
                  quotaUpdateDownstream = TRUE)
state$quotasU[[3]]   # initial quotas
s$quotasU[[3]]       # upstream 1 lost a match
state$quotasD[[3]]
s$quotasD[[3]]       # downstream 1 lost a match
```

## Operation 2: removing agents

The `remove` mode actually takes upstreams and/or downstreams **out** of the
market. There are two variants, controlled by `quotaReset`.

### Slicing mode (`quotaReset = FALSE`, the default)

This drops the listed rows and columns from every per-market array
(`distanceMatrices`, `matchMatrices`, `payoffMatrices`, quotas) and
decrements `noU[m]` / `noD[m]`. The remaining agents are **renumbered**
consecutively from 1.

```{r}
s <- modifyMarket(state, m = 3, u = c(1L, 3L),
                  remove = TRUE, quotaReset = FALSE)
state$noU
s$noU
dim(state$matchMatrices[[3]])
dim(s$matchMatrices[[3]])
```

What was upstream 2 in the original market is now in column 1, and what was
upstream 4 is in column 2:

```{r}
s$matchMatrices[[3]]
```

You can also remove downstreams in the same call:

```{r}
s <- modifyMarket(state, m = 3, u = 1L, d = 4L,
                  remove = TRUE, quotaReset = FALSE)
dim(s$matchMatrices[[3]])
```

### Quota-reset mode (`quotaReset = TRUE`)

Sometimes you want the agents to **stay in the market** but become inactive —
e.g. so you can rematch around them without changing the indexing of any
other variables. `quotaReset = TRUE` zeros their quotas and zeros their
row/column of the match matrix, but leaves all other shapes unchanged.

```{r}
s <- modifyMarket(state, m = 3, u = c(1L, 3L),
                  remove = TRUE, quotaReset = TRUE)
dim(s$matchMatrices[[3]])    # same as before
s$matchMatrices[[3]]         # columns 1 and 3 are zeroed
s$quotasU[[3]]               # corresponding quotas are 0
```

### Cascading quotas with `quotaUpdate`

In either remove mode, `quotaUpdate = TRUE` additionally decrements the
**opposite** stream's quotas to reflect the removed matches. For example, if
the agents you remove were participating in real matches, their downstream
partners now have "free" slots that the rematch step can re-fill.

```{r}
s <- modifyMarket(state, m = 3, u = 1L,
                  remove = TRUE, quotaReset = TRUE, quotaUpdate = TRUE)
state$quotasD[[3]]   # initial downstream quotas in market 3
s$quotasD[[3]]       # decremented at d=1 and d=2 (upstream 1's mates)
```

## Operation 3: rematching

After breaking or removing matches you typically want to re-solve the
assignment problem for the affected market. `rematch = TRUE` does exactly
this by calling `generateAssignmentMatrix()` under the hood.

Rematching requires evaluated payoff matrices in the state. The
`precomp_proof.dat` file used so far has *symbolic* distance attribute
values (`d1111`, `d1112`, ...) — convenient for showing structure, but the
payoff cannot be evaluated on them. We'll switch to `precomp_testdata.dat`,
which is a bundled dataset with numeric attributes, for the remaining
examples.

```{r}
filename2 <- system.file("extdata", "precomp_testdata.dat", package = "maxscoreest")
state2 <- addQuotasFromMatches(importMatched(filename2))
state2$noM
state2$noU
state2$noAttr
```

Compute the payoff matrices for some chosen β (its length must equal
`noAttr - 1`):

```{r}
beta <- c(0.5, 1.0, -0.25, 0.3)
state2$payoffMatrices <- evaluatePayoffMatrices(state2$distanceMatrices, beta)
```

Now combine a removal with a rematch in one call:

```{r}
s <- modifyMarket(state2, m = 1, u = 1L,
                  remove = TRUE, quotaReset = TRUE,
                  quotaUpdate = TRUE, rematch = TRUE)
```

The rematched matrix respects the (decremented) quotas — column 1 is all
zeros because upstream 1's quota was reset to 0, and column/row sums never
exceed `quotasU[[m]]` / `quotasD[[m]]`:

```{r}
all(s$matchMatrices[[1]][, 1] == 0)
all(colSums(s$matchMatrices[[1]]) <= s$quotasU[[1]])
all(rowSums(s$matchMatrices[[1]]) <= s$quotasD[[1]])
```

The mate table is regenerated to match.

You can also rematch without any prior modification, e.g. to see what the
estimated β would predict if the market started from scratch:

```{r}
s <- modifyMarket(state2, m = 1, remove = FALSE, rematch = TRUE)
all(colSums(s$matchMatrices[[1]]) <= s$quotasU[[1]])
```

## Convenience helpers

For the common case of "just slice these agents out", the three small
helpers `removeUpstreams`, `removeDownstreams`, and `removeStreams` are more
concise than the full `modifyMarket` call and accept an additional
`updateMatchMatrix` flag.

```{r}
s <- removeUpstreams(state, m = 1, u = c(1L, 2L))
s$noU
dim(s$matchMatrices[[1]])
```

```{r}
s <- removeDownstreams(state, m = 3, d = c(2L, 4L))
s$noD
dim(s$matchMatrices[[3]])
```

```{r}
s <- removeStreams(state, m = 3, u = 1L, d = 1L)
s$noU
s$noD
```

`updateMatchMatrix = FALSE` slices `distanceMatrices` and the quota vector
but leaves the match matrix and mate untouched. This is useful in the
"precomputed distances" workflow described in the Mathematica module, where
the match matrix is not actively maintained:

```{r}
s <- removeUpstreams(state, m = 3, u = 1L, updateMatchMatrix = FALSE)
dim(s$distanceMatrices[[3]])    # sliced
dim(s$matchMatrices[[3]])       # NOT sliced — caller must rebuild later
```

## Snapshots

`storeState()` returns a deep copy of the state list, including the mate
`data.table` objects (which carry reference semantics in R and would otherwise
be aliased). `restoreState()` is the symmetric operation — useful when you
want to perform several speculative `modifyMarket` calls from a known
checkpoint without re-reading the file each time.

```{r}
snapshot <- storeState(state)

s1 <- modifyMarket(state, m = 3, u = 1L, d = 1L,
                   unmatch = TRUE, remove = FALSE)
s2 <- modifyMarket(restoreState(snapshot), m = 3, u = 2L,
                   remove = TRUE, quotaReset = FALSE)

s1$matchMatrices[[3]][1, 1]   # 0
s2$noU[3]                     # 3
```

In practice, since `modifyMarket` itself snapshots its input before any
modification, you only need explicit `storeState`/`restoreState` if you plan
to mutate the state outside the package's API (for example by editing a list
member directly).

## Helpers for rebuilding mate and quota structures

The two low-level helpers used by `modifyMarket` are also exposed because
they're useful on their own.

`mateFromMatchMatrix()` is the R analogue of Mathematica's `Cmate`. It
rebuilds the mate `data.table` (with columns `UpStream` and `DownMates`) from
a 0/1 match matrix:

```{r}
mateFromMatchMatrix(state$matchMatrices[[3]])
```

`quotaFromMatchMatrix()` is the analogue of Mathematica's `Cquota`. It
returns the per-agent match counts as a list:

```{r}
quotaFromMatchMatrix(state$matchMatrices[[3]])
```

`addQuotasFromMatches()` (shown earlier) is just a list-wide convenience
layer over `quotaFromMatchMatrix()` that adds the `$quotasU` and `$quotasD`
members to a state.

## Correspondence with the Mathematica `modifydata.m`

| Mathematica | R |
| --- | --- |
| `store[var]` / `restore[var]` | `storeState(state)` / `restoreState(snap)` |
| `modify[m, u, d, <\|...\|>]` | `modifyMarket(state, m, u, d, ...)` |
| `removeU[m, u, match]` | `removeUpstreams(state, m, u, updateMatchMatrix)` |
| `removeD[m, d, match]` | `removeDownstreams(state, m, d, updateMatchMatrix)` |
| `removeUD[m, u, d, match]` | `removeStreams(state, m, u, d, updateMatchMatrix)` |
| `Cmate[matchMatrix]` (in matching.m) | `mateFromMatchMatrix(matchMatrix)` |
| `Cquota[matchMatrix]` (in matching.m) | `quotaFromMatchMatrix(matchMatrix)` |

The package's test suite includes a `[parity #1..#6]` group that reproduces
the six Mathematica unmatch-mode tests from
`MSE-Mathematica/testing/modify.testdata` verbatim and compares
`matchMatrix`, `mate`, and `quota` against the exact expected values from the
Mathematica test data file.
