---
title: "Calculating optimal matching from unmatched data"
author: "Panaghis Mavrokefalos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating optimal matching from unmatched data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A quick example

For parameter and index definitions (e.g. `mIdx`, `noAttr`, `dataArray`), see `vignette("glossary", package = "maxscoreest")`.

```{r setup}
library(maxscoreest)
```

First, we need to import the data into R. Let's use some synthetic data provided
with the package.

```{r}
filenameUp <- system.file("extdata", "unmatched_testdata_up.csv", package = "maxscoreest")
filenameDn <- system.file("extdata", "unmatched_testdata_dn.csv", package = "maxscoreest")
```

We import this using the function `importUnmatched`.

```{r}
unmatchedData <- importUnmatched(filenameUp, filenameDn)
```

Next, we need to construct the (unevaluated) payoff matrices from the
attribute matrices of the data.

```{r}
unevalPayoffMatrices <- makeDistanceMatrices(
    unmatchedData$attributeMatricesUp,
    unmatchedData$attributeMatricesDn)
```

Now, we can use a known parameter vector to evaluate them.

---

Note that `beta` is the free parameter vector, and therefore has length one less
than the total number of attributes, which in this case is $3$.

---

```{r}
beta <- c(0.5, 1.5)
payoffMatrices <- evaluatePayoffMatrices(unevalPayoffMatrices, beta)
```

Now we are ready to calculate the optimal matching. We provide the calculated
payoff matrices, as well as the quotas both for upstreams and for downstreams.

```{r}
matchMatrices <- CmatchMatrices(
    payoffMatrices, unmatchedData$quotasUp, unmatchedData$quotasDn)
matchMatrices
```

A full runnable script for this workflow is available as a workbook; see the README (Workbook templates).

# Taking a closer look

## Importing

The `importUnmatched` function can handle delimiter-separated value files with
headers. It supports two modes:

- identifying fields by their position (the default)
- identifying fields by their name

The optional argument `fieldMode` switches between those modes.
The exact format of the header and values is explained in more detail in the
documentation.

## Accessing the imported data

We can see how many markets are in the data, how many members each stream
has in each market, and how many attributes there are.

```{r}
unmatchedData$noM
unmatchedData$noU
unmatchedData$noD
unmatchedData$noAttr
```

We can also view the file headers, containing the field names.

```{r}
unmatchedData$headerUp
unmatchedData$headerDn
```

The `unmatchedData` structure exposes four elements, called
`$attributeMatricesUp`, `$attributeMatricesDn`, `$quotasUp`, and `$quotasDn`.
They are lists of size `noM` each, with each element
containing an array with the relevant data for each market:

```{r}
length(unmatchedData$attributeMatricesUp)
length(unmatchedData$attributeMatricesDn)
length(unmatchedData$quotasUp)
length(unmatchedData$quotasDn)
```

We can view the attribute values for the first upstream of the second market:

```{r}
mIdx <- 2
uIdx <- 1
aIdx <- 3
unmatchedData$attributeMatricesUp[[mIdx]][aIdx, uIdx]
unmatchedData$attributeMatricesUp[[mIdx]][, uIdx]
```

We can also view the quotas for the first market:

```{r}
mIdx <- 1
unmatchedData$quotasUp[[mIdx]]
unmatchedData$quotasDn[[mIdx]]
```

## Accessing the payoff matrices

The structure `payoffMatrices` contains the values of the payoff function,
evaluated for the parameter vector `beta`, for all upstream-downstream pairs.

```{r}
length(payoffMatrices)
mIdx <- 1
dim(payoffMatrices[[mIdx]])
payoffMatrices[[mIdx]]
uIdx <- 1
dIdx <- 2
payoffMatrices[[mIdx]][dIdx, uIdx]
```

## Accessing the match matrices

Like `payoffMatrices`, the structure `matchMatrices` is a list of arrays, with
one element for each market. Each element is an array with elements `0` or `1`.

```{r}
matchMatrices
```

We can check if the upstream-downstream pair $(2, 3)$ of the second market
has a match:

```{r}
mIdx <- 2
uIdx <- 2
dIdx <- 3
matchMatrices[[mIdx]][dIdx, uIdx]
```

We can also view the matching information for a single stream.

```{r}
# Leaving an index blank takes all elements for that dimension.
matchMatrices[[mIdx]][, uIdx]
# `which` returns the indices where the elements above are `1` -- i.e. the
# downstream matches of upstream '2'.
which(matchMatrices[[mIdx]][, uIdx] == 1)
```

